---
title: "Course 5 Class Project 1"
author: "Brent Foster"
date: "January 16, 2018"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

#Import applicable libraries
library("ggplot2")
library("reshape2")
library("gridExtra")

```

## Project: Analysis of personal movement device

#### Loading and preprocessing of the data

```{r dataload}

#Check if file exists.  If it does not, download it and unzip it.
if(!file.exists("repdata%2Fdata%2Factivity.zip")) {
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip","repdata%2Fdata%2Factivity.zip")
  unzip("repdata%2Fdata%2Factivity.zip")
}

#Load data into dataframe
activityDS <- read.csv("activity.csv")

#Aggregate (calculate) total number of steps per day
aggdatatotal <- aggregate(steps ~ date, data = activityDS, FUN=sum,rm=TRUE)

```

#### What is the total number of steps taken per day?
#### Plot, Mean, and Median
#### Note: Missing values were ignored

```{r stepshistogram, echo=TRUE}

#Histogram of the total number of steps taken each day
ggplot(data=aggdatatotal,aes(x=aggdatatotal$date,y=aggdatatotal$steps)) + geom_histogram(stat="identity") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  ggtitle("Total number of steps taken per day") + 
  xlab("Date") + 
  ylab("Steps")

mean(aggdatatotal$steps)
median(aggdatatotal$steps)

```

#### What is the daily activity pattern?
#### Average steps calculation and plot
#### Note: Missing values were ignored

```{r}

#Average number of steps taken across all 5-minute intervals
aggdataavg <- aggregate(steps ~ interval, data = activityDS, FUN=mean,rm=TRUE)

#Change column name to more accurate description
names(aggdataavg)[2] <- paste("avgsteps")

ggplot(data=aggdataavg,aes(x=aggdataavg$interval,y=aggdataavg$avgsteps)) + geom_line(stat="identity") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  ggtitle("Average Steps Taken During Time Interval (Across All Days)") + 
  xlab("5-Second Time Interval") + 
  ylab("Average Steps")

#Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
aggdataavg[which.max(aggdataavg$avgsteps),]


```
####Imputing Missing Values

####Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
#Number of incomplete cases
nrow(activityDS[!complete.cases(activityDS),])
```
####Strategy for filling in missing values:
####Using the mean of the five second interval across all days

```{r}
#Duplicate the original dataset and merge with interval averages
activityDSFI <- merge(activityDS,aggdataavg,by="interval")

#Fill in missing values with the average for the interval
activityDSFI$steps[is.na(activityDSFI$steps)] <- activityDSFI$avgsteps[is.na(activityDSFI$steps)]

#Remove the avgsteps column from dataset to arrive at final version of new dataset
activityDSFI$avgsteps <- NULL
```

#### What is the total number of steps taken per day?
#### Plot, Mean, and Median
#### Note: Missing values were filled in

```{r}
#Aggregate (calculate) total number of steps per day
aggdatatotal2 <- aggregate(steps ~ date, data = activityDSFI, FUN=sum,rm=TRUE)

#Histogram of the total number of steps taken each day
ggplot(data=aggdatatotal2,aes(x=aggdatatotal2$date,y=aggdatatotal2$steps)) + geom_histogram(stat="identity") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  ggtitle("Total number of steps taken per day") + 
  xlab("Date") + 
  ylab("Steps")

mean(aggdatatotal2$steps)

median(aggdatatotal2$steps)
```
####Q:  Do these values differ from the estimates from the first part of the assignment?
####A:  Based on my solution for imputing missing values, the means do not differ but the median's do differ.

####Q:  What is the impact of imputing missing data on the estimates of the total daily number of steps?
####A:  Based on the general trends observed, there appears to be minimal impact on the conclusions obtained.

####The charts below are provided to give a sense of impact (or lack thereof) that the imputed values have on the dataset

```{r}

#Subset the filled in dataset before merging
aggdatatotalMerged <- aggdatatotal

#Merge steps data from two datasets for comparison
aggdatatotalMerged <- merge(aggdatatotalMerged,aggdatatotal2,by="date",all=TRUE)

#Change column name to more accurate description
names(aggdatatotalMerged)[2] <- paste("steps")
names(aggdatatotalMerged)[3] <- paste("stepsFilledIn")

#Histogram of the total number of steps taken each day
plot1 <- ggplot(data=aggdatatotalMerged,aes(x=date,y=steps,group=1)) +
  geom_line(na.rm=TRUE) + geom_point(na.rm=TRUE) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  ggtitle("Total number of steps taken per day (missing values removed)") + 
  xlab("Date") + 
  ylab("Steps")

plot2 <- ggplot(data=aggdatatotalMerged,aes(x=date,y=stepsFilledIn,group=1)) + 
  geom_line() + geom_point() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  ggtitle("Total number of steps taken per day (missing values imputed)") + 
  xlab("Date") + 
  ylab("Steps")

grid.arrange(plot1,plot2,ncol=1)


```


####Are there differences in activity patterns between weekdays and weekends?
####Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)

```{r}
#Create new column that computes the day of the week
activityDSFI$dayType <- ifelse(weekdays(as.Date(activityDSFI$date),abbreviate=TRUE)=='Sat' |
                               weekdays(as.Date(activityDSFI$date),abbreviate=TRUE)=='Sun','weekend','weekday')

#Aggregate (calculate) total number of steps per weekend or weekday day
aggdatatotal3 <- aggregate(list(steps=activityDSFI$steps),by=list(interval=activityDSFI$interval,dayType=activityDSFI$dayType),FUN=mean,rm=TRUE)

ggplot(aggdatatotal3, aes(interval, steps))+geom_line(color="firebrick")+facet_wrap(~dayType,scales="free",nrow=2,ncol=1) +
  theme_bw()
```